\MakeUppercase{The Structure of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-06-18 k-planar.tex
Sending to coauthors

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{pat}
\usepackage{paralist}

@@

% \usepackage{tipa}
% \usepackage{upgreek}
% \usepackage{rotating}
% \newcommand{\V}{\rotatebox[origin=c]{180}{$V$}}
% \newcommand{\Y}{\rotatebox[origin=c]{180}{$Y$}}
%\usepackage{hyperref}

@@

@@

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define. A \emph{tree-decomposition} $\mathcal{K}$ of a graph $G$ consists of a tree $K$ and a collection $\mathcal{K}=\{B_x:x\in V(K)\}$ of subsets of $V(G)$ indexed by nodes of $K$ such that (i)~for every $vw\in E(G)$, there exists some $x\in V(K)$ with $v,w\in B_x$; and (ii)~for every $v\in V(G)$, $K[x] = K[\{y: x\in B_y\}]$ is connected.  The \emph{width} of the tree-decomposition $\mathcal{K}$ is $\max\{|B_x|:x\in V(K)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum integer $t$ such that $G$ has a tree-decomposition of width $t$.

 \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs and (more generally) apex-minor-free graphs.\footnote{A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. \note{PM}{Help me finish this definition of apex-minor-free.}}

In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered width in which $H$ has bounded treewidth: $k$-planar graphs, which we now define. An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve in a surface $S$ is a continuous function $f:[0,1]\to S$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists $k$-plane graph $G$ isomorphic to $G'$.

Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs. Thus, like genus-$g$ graphs,\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} $k$-planar graphs are a generalization of planar graphs (which are genus-$0$ graphs).  Unlike genus-$g$ graphs, however, the family of $k$-planar graphs is not minor-closed.\footnote{A class $\mathcal{G}$ of graphs is \emph{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \emph{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class.}  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors.

@@

@@

The purpose of this section is to prove \thmref{k-planar}.  In this section, all graphs are finite, simple, and undirected.  For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$) we use $G[S]$ to denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$ and for a graph $G'$, we use $G-G'$ as a shorthand for $G-V(G')$.  We use $G'\subseteq G$ to denote subgraph containment, i.e., $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$. \note{PM}{TODO: Hunt down occurrences of $G-G'$ and replace with $G-V(G')$.}

We begin by introducing a structural tool for planar graphs, called a PS-tree, that is implicit in previous work on $H$-partitions of planar graphs \cite{dujmovic.joret.ea:planar}.  We then show how the PS-tree of a planarized version of a $k$-planar graph $G$ can be used to find a layered $H$-partition of $G$ of small layered width in which $H$ has small treewidth.

@@

Let $T$ be a tree rooted at some node $r\in V(T)$. The \emph{$T$-depth} of a node $x\in V(T)$ is equal to the length of the path from $x$ to $r$ in $T$. A path $P$ in $T$ is \emph{vertical} if, for each integer $i$, $P$ contains at most one vertex of $T$-depth $i$.  The deepest vertex in a vertical path $P$ is called  the \emph{lower endpoint} of $P$ and the shallowest vertex in $P$ is called the \emph{upper endpoint} of $P$. For each node $v\in V(T)$, we call each node on the path, in $T$, from $v$ to $r$ (including $v$ itself) a \emph{$T$-ancestor} of $v$.  If $v'$ is a $K$-ancestor of $v$, then $v$ is a \emph{$K$-descendant} of $v'$.

A \emph{near-triangulation} $N$ is a plane graph whose outer face, $F$, is bounded by a simple cycle in $N$ and each of whose inner faces is bounded by a 3-cycle in $N$.  A \emph{triangulation} is a near-triangulation whose outer face is bounded by a 3-cycle.  For any cycle $F$ in a near-triangulation $N$ we use $N\circledast F$ to denote the near-triangulation consisting of all the edges and vertices of $N$ that are contained in $F$ or its interior.

For a connected graph $G$ and two vertices $v,w\in V(G)$, we use $\dist_{G}(v,w)$ to denote the length of a shortest path, in $G$, with endpoints $v$ and $w$.  For a vertex $r\in V(G)$, the \emph{BFS layering} of $G$ from $r$ is the layering $\langle V_0,V_1,\ldots\rangle$ of $G$ where $V_i=\{v\in V(G): \dist_{G}(r,v)=i\}$ for each integer $i\ge 0$.  A \emph{BFS spanning tree} $T$ of $G$ rooted at $r$ is a spanning tree of $G$ rooted at $r$ where, for every $v\in V(G)\setminus\{r\}$, the parent $v'$ of $v$ in $T$ has $\dist_G(r,v')= \dist_G(r,v)-1$.

@@

Fix some cycle $F$ in $\Delta$ and let $N=\Delta\circledast F$.
Let $\tau=v_1v_2v_3$ be an inner face of $N$ and consider the three minimal vertical paths $Q_1$, $Q_2$, and $Q_3$ in $T$ such that, for each $i\in\{1,2,3\}$ the lower endpoint of $Q_i$ is $v_i$ and the upper endpoint $v_i'$ of $Q_i$ is in $V(F)$.  If $Q_1$, $Q_2$, and $Q_3$ are vertex disjoint, then the graph $\overline{Y}=\tau\cup Q_1\cup Q_2\cup Q_3$ is called the \emph{closed tripod} in $N$ determined by $\tau$.  We call $Q_1$, $Q_2$, and $Q_3$ the \emph{legs}, $v_1'$, $v_2'$ and $v_3'$ the \emph{feet}, and $\tau$ the \emph{crotch} of $\overline{Y}$.  We call $Y=\overline{Y}-F$ the \emph{(open) tripod} of $N$ determined by $\tau$.  The graph $M=\overline{Y}\cup F$ is called a \emph{Mercedes graph} and the (at most three) inner faces of $M$ other than $\tau$ are called the \emph{wedges} of $M$.

Let $r$ be a degree-3 node on the outer face of a triangulation $\Delta$.  A \emph{PS-tree}  $K=K(\Delta, T, r)$ is a rooted tree in which each non-root node $x\in V(K)$ is associated with several objects (refer to \figref{K-node}):

@@

\begin{enumerate}
  \item A cycle $F_x$ in $\Delta$ and the near-triangulation $N_x=\Delta\circledast F_x$.

  \item An inner face $\tau_x$ of $N_x$ that determines a closed tripod $\overline{Y}_x$ and an open tripod $Y_x=\overline{Y}_x-F_x$ in $N_x$.

  \item The \emph{Mercedes graph} $M_x=F_x\cup \overline{Y}_x$
\end{enumerate}

The root node $r^0_K$ of $K$ is exceptional.  We define $F_{r^0_K}$ to be the single-vertex graph containing $r$, we define $\overline{Y}_{r^0_K}$ to be the star formed by $r$ and its three neighbours in $\Delta$, and (as usual) $Y_{r^0_K}=\overline{Y}_{r^0_K}-F_{r^0_K}$.  The root node $r^0_K$ has exactly one child $r_K$, that we call the \emph{sub-root} of $K$, for which $F_{r_K}$ is the 3-cycle formed by the neighbours of $r$ in $\Delta$.

The non-root nodes of the PS-tree $K$ (including the sub-root $r_K$) satisfy the following conditions:
\begin{enumerate}[(PR1)]
  \item \proplabel{children} For each non-root node $x\in V(K)$ such that $M_x$ has $f\le 3$ wedges $F_{x,1},\ldots,F_{x,f}$, the node $x$ has exactly $f$ children $y_1,\ldots,y_f$ in $K$ and $F_{y_i} = F_{x,i}$ for each $i\in\{1,\ldots,f\}$.

@@

For any triangulation $\Delta$ in which the outer face has a vertex $r$ of degree 3 and for any BFS spanning tree of $\Delta$ rooted at $r$, there exists a PS-tree $K=K(\Delta,T,r)$.  Indeed, this tree is constructed (albeit implicitly) by \citet[Proof of Lemma~14]{dujmovic.joret.ea:planar}.\footnote{\citet{dujmovic.joret.ea:planar} do not require that $T$ be a BFS spanning tree, only that its root, $r$ have degree 3 in $\Delta$ and in $T$.}

@@

Let $r$ be the vertex of the $K_4$ not incident on $F_0$ and let $T$ be a BFS spanning tree of $G^{++}$ rooted at $r$. 
For each integer $i\ge 0$, let $V_i=\{v\in V(G^+): \dist_{T}(r,v)=i\}$ and observe that $\{V_2,V_3,\ldots,V_{h}\}$ is a layering of $G^+$ (but not of $G$).

Let $K=K(G^{++},T,r)$ be a PS-tree and observe that the sub-root $r_K$ of $K$ has $F_{r_K}=F_0$.  This implies:
\begin{compactenum}[(PR1)]\setcounter{enumi}{4}
  \item \proplabel{all-in}  $G^+\subseteq N_{r_K}-F_{r_K}$.
\end{compactenum}

By \propref{all-in}, $P_{z}$ contains the sub-root $r_K$ of $K$.  By \propref{children}, any node $x\in P_z$ has at most one of its children in $P_z$ and any node $x'\not\in P_z$ has none of its children in $P_z$.  Therefore $K[P_z]$ is a vertical path that contains $r_K$ (and whose lower endpoint is the unique node $x\in V(K)$ such that $z\in Y_x$). We are about to repeatedly use the fact that the common intersection of two or more vertical paths is a vertical path.

For each edge $vw\in E(G)$, consider the walk $W_{vw}$ in $G^+$ that corresponds to the sequence of vertices in $G^+$ encountered while walking along the curve $vw$ from $v$ to $w$. We define $A_{vw}=\bigcap_{z\in W_{vw}} P_z$.  Since $V(W_{vw})\subseteq V(G^+)$, $A_{vw}$ is well-defined.  For each $z\in V(W_{vw})$, $K[P_{z}]$ is a vertical path in $K$ that contains $r_K$.   Therefore $K[A_{vw}]$ is a vertical path in $K$ that contains $r_K$.  For each vertex $v\in V(G)$ define $A_v=\bigcap_{vw\in E(G)} A_{vw}$.  Again, $K[A_v]$ is a vertical path in $K$ that contains $r_K$.  We define $a(v)$ to be the lower endpoint of $K[A_v]$.  In words, $a(v)$ is the lowest node in $K$ such that the interior of $F_{a(v)}$ contains all the edges of $G$ incident to $v$.

For each node $x\in V(K)$, define the \emph{separator}
\[
   S_x = \{v\in V(G): a(v)=x \} \enspace .
\]
Clearly $\mathcal{S}=\{S_x:x\in V(K)\}$ is a partition of $V(G)$.  We will use $\mathcal{S}$ as the partition that defines our quotient graph $H=G/\mathcal{S}$ with vertex set $V(H)=V(K)$ and the edge $xy\in E(H)$ if and only if there exists an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_y$.  First we argue that this partition has small layered width with respect to the layering $\{V_2,V_3,\ldots\}$ of $G^+$ defined by $T$ (recall that this is a layering of $G^+$, not a layering of $G$).

@@

\begin{proof}
  Since $v\in S_x$, the lower endpoint of $K[A_{vw}]$ is $x$ for some edge $vw\in E(G)$.  This implies that $W_{vw}\subseteq N_x-F_x$ but $W_{vw} \not\subseteq N_{x'}-F_{x'}$ for every child $x'$ of $x$ in $K$. Therefore, \propref{children} and \propref{ancestor-boundary} imply that $W_{vw}$ contains a vertex $v'\in Y_x$.  Since $G$ is $k$-planar, the length of $W_{vw}$ is at most $k+1$.  Therefore, $\dist_{G^{++}}(v,v') \le k+1$.
  By the triangle inequality,     
  \[
    \dist_{G^{++}}(r,v') \le \dist_{G^{++}}(r,v) + \dist_{G^{++}}(v,v') 
       \le \dist_{G^{++}}(r,v) + k + 1 \enspace .
  \]
  Similarly, $\dist_{G^{++}}(r,v) \le \dist_{G^{++}}(r, v') + k+1$.  Therefore
  $|\dist_{G^+}(r,v)-\dist_{G^+}(r,v')| \le k+1$ which establishes the claim, since $T$ is BFS spanning tree of $G^{++}$ rooted at $r$, so $\dist_{T}(r,z)=\dist_{G^{++}}(r,z)$ for all $z\in V(G^+)$.
\end{proof}

\begin{clm}\clmlabel{width-of-g-plus}
  For any $x\in V(H)$, and any integer $i\ge 0$, $|S_x\cap V_i|\le 24k+12$.
\end{clm}

\begin{proof}
  Since the vertices of $Y_x$ come from at most three vertical paths in $T$, $|V(Y_x)\cap V_i|\le 3$.  \clmref{chargeback-to-y} shows that, for any $v\in S_x\cap V_i$, there exists $vw\in E(G)$ and $v'\in Y_x\cap V(W_{vw})$ with $\dist_T(v,v')\le k+1$. Therefore, $v'\in V(Y_x)\cap V_{j}$ for some $j\in\{i-k-1,\ldots,i+k+1\}$.  If $v'=v$ (so $v'\in V_i$) then $v'$ contributes only one vertex to $S_x$.  If $v'\neq v$ then $v'\in V(G^+-G)$ is a point where two edges of $G$ cross and $v'$ contributes at most the four endpoints of these edges to $S_x$. In either case each $v'\in V(Y_x)\cap \bigcup_{j=i-k-1}^{i+k+1} V_j$ contributes at most four vertices to $S_x\cap V_i$, so $|S_x\cap V_i|\le 3\times 4\times (2k+3) = 24k+36$.
\end{proof}

>>>>

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}

@@

@@

In 1965, \citet{ringel:sechsfarbenproblem} introduced the notion of $1$-planar graphs, a generalization of planar graphs that allows edges to cross provided that no edge is involved in more than one crossing. This simple generalization has since become a rich source of problems and results.  Indeed, the annotated bibliography on 1-planar graphs by \cite{kobourov.liotta.ea:annotated} contains more than 140 entries.

The natural generalization of $1$-planar graphs, $k$-planar graphs (that allow edge to be involved in up to $k$ crossings), has been the topic of important recent work on graph drawing. Though planar graphs are, by now, quite well-understood, the situation is much less clear for $k$-planar graphs; even the maximum edge-density of $k$-planar graphs is not completely settled.  For $k\le 4$, the maximum number of edges in an $n$-vertex $k$-planar graph is $(k+3)(n-2)$, which is tight for $k=1,2$ \cite{pach.toth:graphs}.  A near-tight bound of $11(n-2)/2$ is known for $k=3$ \cite{pach.radoicic.ea:improving} but for $k>4$, the current best upper bound is $4.108\sqrt{k}n$ \cite{pach.toth:graphs}.

% In the current paper we investigate the structure of $k$-planar graphs.  Since planarity is closed under taking minors, planar graphs can be characterized by a finite set of forbidden minors.  Indeed, Wagner's Theorem states that this set consists of the complete 5-vertex graph $K_5$ and the complete bipartite graph $K_{3,3}$.  Unfortunately, $k$-planarity is not closed under taking minors, ruling out this kind of clean structural result.  The purpose of the current paper is to extend a new structural result for planar graphs to $k$-planar graphs.

The purpose of the current paper is to show that new structural results for planar graphs can be extended to $k$-planar graphs.  These results can be used to give a rough characterization of $k$-planar graphs of powers of bounded-degree planar graphs
% Rather, we present a rough structural characterization of $k$-planar graphs that, despite its roughness, provides several new results for $k$-planar and related graph families.  
All of these results are based on layered $H$-partitions, which we now define.
{\color{red}TODO: Help.}

@@

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define. A \emph{tree-decomposition} $\mathcal{K}$ of a graph $G$ consists of a tree $K$ and a collection $\mathcal{K}=(B_x:x\in V(K))$ of subsets of $V(G)$ indexed by nodes of $K$ such that (i)~for every $vw\in E(G)$, there exists some $x\in V(K)$ with $v,w\in B_x$; and (ii)~for every $v\in V(G)$, the induced subgraph $K[x] := K[\{y: x\in B_y\}]$ is connected.  The \emph{width} of the tree-decomposition $\mathcal{K}$ is $\max\{|B_x|:x\in V(K)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.

 \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs, graphs of bounded genus, and (more generally) apex-minor-free graphs.\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \emph{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \emph{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class.

 A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \emph{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}

% In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered width in which $H$ has bounded treewidth:
In the current paper we show that $k$-planar graphs also admit layered $H$-partitions in which $H$ has bounded treewidth. First, we formally define $k$-planar graphs.  An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists $k$-plane graph $G$ isomorphic to $G'$.

Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs. Thus, like genus-$g$ graphs,\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} $k$-planar graphs are a generalization of planar graphs (which are genus-$0$ graphs).  Unlike genus-$g$ graphs, however, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors.

@@

\note{PM}{TODO: Write here about applications. non-repetitive colouring, $p$-centered colouring, $(g,k)$-planar graph, $(g,d)$-map graphs, $k$-NN graphs, David's rough characterization.}

@@

The purpose of this section is to prove \thmref{k-planar}.  In this paper, all graphs are finite and undirected. Unless specifically mentioned otherwise, all graphs are also simple.  For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$) we use $G[S]$ to denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$ and for a graph $G'$, we use $G-G'$ as a shorthand for $G-V(G')$.  We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$.

We begin by introducing a structural tool for planar graphs, called a tripod decomposition, that is implicit in previous work on $H$-partitions of planar graphs \cite{dujmovic.joret.ea:planar}.  We then show how the tripod decomposition of a planarized version of a $k$-planar graph $G$ can be used to find a layered $H$-partition of $G$ of small layered width in which $H$ has small treewidth.

@@

Let $T$ be a tree rooted at some node $r\in V(T)$. The \emph{$T$-depth} of a node $x\in V(T)$ is equal to the length of the path from $x$ to $r$ in $T$. A path $P$ in $T$ is \emph{vertical} if, for each integer $i$, $P$ contains at most one vertex of $T$-depth $i$.  The deepest vertex in a vertical path $P$ is called  the \emph{lower endpoint} of $P$ and the shallowest vertex in $P$ is called the \emph{upper endpoint} of $P$. For each node $v\in V(T)$, we call each node on the path, in $T$, from $v$ to $r$ (including $v$ itself) a \emph{$T$-ancestor} of $v$.  If $v'$ is a $T$-ancestor of $v$, then $v$ is a \emph{$T$-descendant} of $v'$.

A \emph{near-triangulation} $N$ is a plane graph whose outer face, $F$, is bounded by a simple cycle in $N$ and each of whose inner faces is bounded by a 3-cycle in $N$.  A \emph{triangulation} is a near-triangulation whose outer face is bounded by a 3-cycle.  For any cycle $F$ in a near-triangulation $N$ we use $N\llparenthesis F\rrparenthesis$ to denote the near-triangulation consisting of all the edges and vertices of $N$ that are contained in $F$ or its interior.

For a connected graph $G$ and two vertices $v,w\in V(G)$, we use $\dist_{G}(v,w)$ to denote the length of a shortest path, in $G$, with endpoints $v$ and $w$.  For a vertex $r\in V(G)$, the \emph{BFS layering} of $G$ from $r$ is the layering $\langle V_0,V_1,\ldots\rangle$ of $G$ where $V_i=\{v\in V(G): \dist_{G}(r,v)=i\}$ for each integer $i\ge 0$.  A \emph{BFS spanning tree} $T$ of $G$ rooted at $r$ is a spanning tree of $G$ rooted at $r$ where, for every $v\in V(G)$, $\dist_T(v,r)=\dist_G(v,r)$.

@@

Fix some cycle $F$ in $\Delta$ and let $N=\Delta\llparenthesis F\rrparenthesis$.
Let $\tau=v_1v_2v_3$ be an inner face of $N$ and consider the three minimal vertical paths $Q_1$, $Q_2$, and $Q_3$ in $T$ such that, for each $i\in\{1,2,3\}$ the lower endpoint of $Q_i$ is $v_i$ and the upper endpoint $v_i'$ of $Q_i$ is in $V(F)$.  If $Q_1$, $Q_2$, and $Q_3$ are vertex-disjoint, then the graph $\overline{Y}:=\tau\cup Q_1\cup Q_2\cup Q_3$ is called the \emph{closed tripod} in $N$ determined by $\tau$.  We call $Q_1$, $Q_2$, and $Q_3$ the \emph{legs}, $v_1'$, $v_2'$ and $v_3'$ the \emph{feet}, and $\tau$ the \emph{crotch} of $\overline{Y}$.  We call $Y:=\overline{Y}-F$ the \emph{(open) tripod} of $N$ determined by $\tau$.  The graph $M:=\overline{Y}\cup F$ is called a \emph{Mercedes graph} and the (at most three) inner faces of $M$ other than $\tau$ are called the \emph{wedges} of $M$.

Let $r$ be a degree-3 node on the outer face of a triangulation $\Delta$.  A \emph{tripod decomposition}  $K=K(\Delta, T, r)$ is a rooted tree in which each non-root node $x\in V(K)$ is associated with several objects (refer to \figref{K-node}):

@@

\begin{enumerate}
  \item A cycle $F_x$ in $\Delta$ and the near-triangulation $N_x:=\Delta\llparenthesis F_x\rrparenthesis$.

  \item An inner face $\tau_x$ of $N_x$ that determines a closed tripod $\overline{Y}_x$ and an open tripod $Y_x:=\overline{Y}_x-F_x$ in $N_x$, both with crotch $\tau_x$.

  \item The \emph{Mercedes graph} $M_x:=F_x\cup \overline{Y}_x$
\end{enumerate}

The root node $r^0_K$ of $K$ is exceptional.  We define $F_{r^0_K}$ to be the single-vertex graph containing $r$, we define $\overline{Y}_{r^0_K}$ to be the star formed by $r$ and its three neighbours in $\Delta$, and (as usual) $Y_{r^0_K}:=\overline{Y}_{r^0_K}-F_{r^0_K}$.  The root node $r^0_K$ has exactly one child $r_K$, called the \emph{sub-root} of $K$, for which $F_{r_K}$ is the 3-cycle formed by the neighbours of $r$ in $\Delta$.

The non-root nodes of the tripod decomposition $K$ (including the sub-root $r_K$) satisfy the following conditions:
\begin{enumerate}[(PR1)]
  \item \proplabel{children} For each non-root node $x\in V(K)$ such that $M_x$ has $f\le 3$ wedges $F_{x,1},\ldots,F_{x,f}$, the node $x$ has exactly $f$ children $y_1,\ldots,y_f$ in $K$ and $F_{y_i} = F_{x,i}$ for each $i\in\{1,\ldots,f\}$.

@@

For any triangulation $\Delta$ in which the outer face has a vertex $r$ of degree 3 and for any BFS spanning tree of $\Delta$ rooted at $r$, there exists a tripod decomposition $K=K(\Delta,T,r)$.  Indeed, this tree is constructed (albeit implicitly) by \citet[Proof of Lemma~14]{dujmovic.joret.ea:planar}.\footnote{\citet{dujmovic.joret.ea:planar} do not require that $T$ be a BFS spanning tree, only that its root, $r$ has degree 3 in $\Delta$ and in $T$.}

@@

Let $r$ be the vertex of the $K_4$ not incident on $F_0$ and let $T$ be a BFS spanning tree of $G^{++}$ rooted at $r$. 
For each integer $i\ge 0$, let $V_i:=\{v\in V(G^+): \dist_{T}(r,v)=i\}$ and observe that $\{V_2,V_3,\ldots,V_{h}\}$ is a layering of $G^+$ (but not of $G$).

Let $K=K(G^{++},T,r)$ be a tripod decomposition and observe that the sub-root $r_K$ of $K$ has $F_{r_K}=F_0$.  This implies:
\begin{compactenum}[(PR1)]\setcounter{enumi}{4}
  \item \proplabel{all-in}  $G^+\subseteq N_{r_K}-F_{r_K}$.
\end{compactenum}

For each vertex $z\in V(G^+)$, let $P_z=\{a\in V(K): z\in V(N_a-F_a)\}$.  By \propref{all-in}, $P_{z}$ contains the sub-root $r_K$ of $K$.  By \propref{children}, any node $x\in P_z$ has at most one of its children in $P_z$ and any node $x'\not\in P_z$ has none of its children in $P_z$.  Therefore $K[P_z]$ is a vertical path in $K$ that contains $r_K$ (and whose lower endpoint is the unique node $x\in V(K)$ such that $z\in Y_x$). We are about to repeatedly use the fact that the common intersection of two or more vertical paths is a vertical path.

For each edge $vw\in E(G)$, consider the walk $W_{vw}$ in $G^+$ that corresponds to the sequence of vertices in $G^+$ encountered while walking along the curve $vw$ from $v$ to $w$. We define $A_{vw}:=\bigcap_{z\in W_{vw}} P_z$.  Since $V(W_{vw})\subseteq V(G^+)$, $A_{vw}$ is well-defined.  For each $z\in V(W_{vw})$, $K[P_{z}]$ is a vertical path in $K$ that contains $r_K$.   Therefore $K[A_{vw}]$ is a vertical path in $K$ that contains $r_K$.  For each vertex $v\in V(G)$ define $A_v:=\bigcap_{vw\in E(G)} A_{vw}$.  Again, $K[A_v]$ is a vertical path in $K$ that contains $r_K$.  We define $a(v)$ to be the lower endpoint of $K[A_v]$.  In words, $a(v)$ is the lowest node in $K$ such that the interior of $F_{a(v)}$ contains all the edges of $G$ incident to $v$.

For each node $x\in V(K)$, define the \emph{separator}
\[
   S_x := \{v\in V(G): a(v)=x \} \enspace .
\]
Clearly $\mathcal{S}:=\{S_x:x\in V(K)\}$ is a partition of $V(G)$.  We will use $\mathcal{S}$ as the partition that defines our quotient graph $H:=G/\mathcal{S}$ with vertex set $V(H):=V(K)$ and the edge $xy\in E(H)$ if and only if there exists an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_y$.  First we argue that this partition has small layered width with respect to the layering $\{V_2,V_3,\ldots\}$ of $G^+$ defined by $T$.
 % (recall that this is a layering of $G^+$, not a layering of $G$).

@@

\begin{proof}
  Since $v\in S_x$, the lower endpoint of $K[A_{vw}]$ is $x$ for some edge $vw\in E(G)$.  This implies that $W_{vw}\subseteq N_x-F_x$ but $W_{vw} \not\subseteq N_{x'}-F_{x'}$ for every child $x'$ of $x$ in $K$. Therefore, \propref{children} and \propref{ancestor-boundary} imply that $W_{vw}$ contains a vertex $v'\in Y_x$.  Since $G$ is $k$-planar, the length of $W_{vw}$ is at most $k+1$.  Therefore, $\dist_{G^{++}}(v,v') \le k+1$.
  By the triangle inequality,     
  \[
    \dist_{G^{++}}(r,v') \le \dist_{G^{++}}(r,v) + \dist_{G^{++}}(v,v') 
       \le \dist_{G^{++}}(r,v) + k + 1 \enspace .
  \]
  Similarly, $\dist_{G^{++}}(r,v) \le \dist_{G^{++}}(r, v') + k+1$.  Therefore
  $|\dist_{G^{++}}(r,v)-\dist_{G^{++}}(r,v')| \le k+1$ which establishes the claim, since $T$ is BFS spanning tree of $G^{++}$ rooted at $r$, so $\dist_{T}(r,z)=\dist_{G^{++}}(r,z)$ for all $z\in V(G^+)$.
\end{proof}

\begin{clm}\clmlabel{width-of-g-plus}
  For any $x\in V(H)$, and any integer $i\ge 0$, $|S_x\cap V_i|\le 24k+36$.
\end{clm}

\begin{proof}
  Since the vertices of $Y_x$ come from at most three vertical paths in $T$, $|V(Y_x)\cap V_i|\le 3$.  \clmref{chargeback-to-y} shows that, for any $v\in S_x\cap V_i$, there exists $vw\in E(G)$ and $v'\in Y_x\cap V(W_{vw})$ with $\dist_T(v,v')\le k+1$. Therefore, $v'\in V(Y_x)\cap V_{j}$ for some $j\in\{i-k-1,\ldots,i+k+1\}$.  If $v'=v$ (so $v'\in V_i$) then $v'$ contributes only one vertex to $S_x$.  If $v'\neq v$ then $v'\in V(G^+-G)$ is a point where two edges of $G$ cross and $v'$ contributes at most the four endpoints of these edges to $S_x$. In either case, each $v'\in V(Y_x)\cap \bigcup_{j=i-k-1}^{i+k+1} V_j$ contributes at most four vertices to $S_x\cap V_i$, so $|S_x\cap V_i|\le 3\times 4\times (2k+3) = 24k+36$.
\end{proof}
